---
title: "Movie Recommendation with Market Basket Analysis"
output: 
    html_document:
        toc: true
---

# Introduction
In this project, we applied a data mining algorithm, Apriori, to mine a relationship among films and build a movie recommendation engine. Apriori is a technique in Market Basket Analysis used to discover items that are frequently sold together. Frequently purchased itemset suggests marketing opportunity when customers displayed interest in the subset items. In this case, movies can be viewed as a set of items. We obtained our training data from MovieLens's website(<http://grouplens.org/datasets/movielens/>). We used `MovieLens 20M Dataset` dataset which consisted of 20,000,263 user ratings, across 27,278 movies and 138,493 raters. We found that the mining technique can be utilized to uncover an underlying connection within the movies. It can also be used in a movie recommendation, but a number of suggested films can be quite limited and the quality of such suggestions can be vary.

```{r setup, include=FALSE}
#cache the result
knitr::opts_chunk$set(cache=TRUE)
```

# Explore Movie Data
We extract the compressed file in `ml-20m` directory. We interest in 2 data files:  

* movies.csv(contains information about movies)
* ratings.csv(user ratings)

We firstly investigate movies.csv file

```{r load_package,message=FALSE}
#load all required package
library(arules)
library(dplyr)
library(reshape2)
library(Matrix)
library(stringr)
library(stringdist)
library(ggplot2)

setwd("/home/vitidn/mydata/repo_git/MovieMBA/")
```

```{r load_data}
movies = read.csv("ml-20m/movies.csv", 
                  colClasses = c("integer","character","character"),
                  sep = ",",
                  stringsAsFactors = FALSE)

head(movies)
```

We separate released year from a title

```{r}
movies$year = as.numeric(str_sub( str_trim(movies$title) ,start = -5,end = -2))
```

Some movies have no release year/invalid title format. We note their movieId and discard them for this analysis.
```{r}
discard_movie_id = which(is.na(movies$year))
#display discarded movies
movies$title[discard_movie_id]

movies = movies[-discard_movie_id,]
```

Title is extracted

```{r}
movies$title = str_sub( str_trim(movies$title) ,start = 1,end = -8)
```

Next, we would like to extract genres for each movie(noted that each film can belong to more than one genre). We look at total number of genres.

```{r}
all_genres = unique(unlist(str_split(movies$genres,"\\|")))

all_genres
```

We see 2 genres that are really not a genre definition. We investigate a number of movies without genre defined.

```{r}
movies %>% filter(str_detect(genres,"(no genres listed)") ) %>% nrow()
```

We create binary dummy variables for another 18 genres. We assign each film to the genre it belongs. We discard "IMAX" genre and assign every genres to movies without genre identified. We check the transformed result and drop `genres` column.

```{r}
all_genres = all_genres[! all_genres %in% c("IMAX","(no genres listed)")]

for(genre in all_genres){
    movies[str_c("genre_",genre)] = ifelse(( str_detect(movies$genres,genre) | str_detect(movies$genres,"no genres") ) , 1 , 0)
}

#check the result
head(movies)
tail(movies)

movies$genres = NULL
```

We explore a number of movies for each year in the dataset that we have

```{r}
ggplot(movies,aes(x=year)) + geom_bar() + ggtitle("Number of Movies")
```

We also explore a distributon of each movie genres

```{r}
genre_dist = colSums(movies[,4:21])
genre_dist_df = data.frame(genre = names(genre_dist),count = genre_dist)
genre_dist_df$genre = factor(genre_dist_df$genre,levels = names(sort(genre_dist,decreasing = FALSE)))

ggplot(genre_dist_df,aes(x=genre,y=count,fill=genre)) + 
    geom_bar(stat = "identity") + 
    coord_flip() + 
    ggtitle("Genre Distribution") + 
    theme(legend.position = "none")
```

Now, we get a basic understanding of our movie dataset. Genre and year that we extracted will served as a filter that users can use to narrow down their interest.

# Construct Association Rules from Rating Data

We proceed to read ratings.csv and investigate the dataset. We skip reading `rating` and `timestamp` columns. Noted that we ignore the actual rating here as we put more focus on the fact that the scored movies hold some interesting quality that they at least led the viewers to view them.

```{r}
ratings = read.csv("ml-20m/ratings.csv",
                   colClasses = c("integer","integer","NULL","NULL"),
                   sep=",",
                   stringsAsFactors = FALSE)

head(ratings)
```

We discard ratings that contain id in `discard_movie_id`
```{r}
ratings = ratings %>% filter(! movieId %in% discard_movie_id )
```

We look at a total number of ratings left

```{r}
dim(ratings)[1]
```

We use `arules` package to perform the frequent itemset mining with Apriori algorithm. We construct User-Item matrix with binary values; 0 - a movie isn't seen by a user, and 1 - it is seen. The package use a sparse matrix object, `transactions`, to represent User-Item matrix. This prevents our computing machine from consuming all available RAM as most elements in the matrix will be zero.

```{r}
#convert rating-per-row dataframe into sparse User-Item matrix
user_item_matrix <- as(split(ratings[,"movieId"], ratings[,"userId"]), "transactions")

#investigate the User-Item matrix
#transactions (rows) -> number of raters
#items (columns) -> number of movies
user_item_matrix
```

```{r echo=FALSE}
#clear unused variable
rm(ratings)
gc()
```

Next, we mine for a frequent pair of movies that raters watched. We hypothesize that if movie A and B are frequently viewed together, there should be some underlying relationships between them that incite viewer's curiosity. We can use such finding to recommend movie B to a user if he/she already saw A(or vice versa).  

We set the support threshold to 0.001(the pair is watched together by at least 139 raters) and the minimum confidence(the likelihood that if user watched movie A, he/she will also watch movie B ) to 70%.

```{r}
rule_param = list(
    supp = 0.001,
    conf = 0.7,
    maxlen = 2
)
```

We run Apriori based on the specified rule

```{r}
assoc_rules = apriori(user_item_matrix,parameter = rule_param)
```

We summarize the association rule

```{r}
summary(assoc_rules)
```

We constructed 189611 rules here. We also get summary statistics of "lift" for all rules. Lift is used to measure how the rule "if an user watched A then he will proceed to watch B" performs against chance. For example, if movie B is watched by every users, then the rule A => B will have 100% confidence but this rule will not be really interesting as there is no point to recommend it because everyone tend to watch it anyway. We can use lift to filter the "interestingness" of each rule. Lift equal 1 suggests that A and B are independent. The higher the number, the more they related.

With such huge number of rules, we filter only those that have lift exceed their 75% percentile(4.323).

```{r}
assoc_rules = subset(assoc_rules, lift >= 4.323)

summary(assoc_rules)
```

We cast `assoc_rules` to data.frame and look at some of the data

```{r}
assoc_rules = as(assoc_rules,"data.frame")

head(assoc_rules)
```

The rules still contain `movieId`. We split movies in both sides to a new column

```{r}
rules = sapply(assoc_rules$rules,function(x){
    x = gsub("[\\{\\}]", "", regmatches(x, gregexpr("\\{.*\\}", x))[[1]])
    x = gsub("=>",",",x)
    x = str_replace_all(x," ","")
    return( x )
})

rules = as.character(rules)
rules = str_split(rules,",")

assoc_rules$lhs_movie = sapply( rules, "[[", 1)
assoc_rules$rhs_movie = sapply( rules , "[[", 2)

assoc_rules$rules = NULL
rm(rules)
gc()

assoc_rules$lhs_movie = as.numeric(assoc_rules$lhs_movie)
assoc_rules$rhs_movie = as.numeric(assoc_rules$rhs_movie)
```

We join `assoc_rules` with `movies` to get titles on the left hand side and right hand side of the rule, and also their corresponding genres and released year.

```{r}
assoc_rules = assoc_rules %>% left_join(movies,by=c("lhs_movie" = "movieId") )

assoc_rules$lhs_movie = NULL
col_name = colnames(assoc_rules)
col_name[5:24] = str_c("left.",col_name[5:24])
colnames(assoc_rules) = col_name

assoc_rules = assoc_rules %>% left_join(movies,by=c("rhs_movie" = "movieId"))
assoc_rules$rhs_movie = NULL
col_name = colnames(assoc_rules)
col_name[24:43] = str_c("right.",col_name[24:43])
colnames(assoc_rules) = col_name
```

```{r echo=FALSE}
rm(movies)
gc()
```

# Mining the Relationship and Recommending Movies

Now, we can look at the rules we mined. For example, we can look at top rules with highest lift.

```{r}
assoc_rules %>% arrange(desc(lift)) %>% select(left.title,left.year,right.title,right.year,support,confidence,lift) %>% head()
```

For the top rules, we discover sequel/prequel relationship between the movies. We would like to find recommendations that have not-so-obvious relationship instead.

We can filter out results that have sequel-prequel relationship based on their similar titles. We do a naive filter here. Result with number on both sides or similar opening string is removed, we also exclude the "Thin man" serie.

```{r}
assoc_rules = assoc_rules %>% 
    filter( ! (grepl("[0-9]",left.title,perl = TRUE) &  grepl("[0-9]",right.title,perl = TRUE) ) ) %>%
    filter( ! (grepl("Thin Man",left.title,perl = TRUE) &  grepl("Thin Man",right.title,perl = TRUE) ) ) %>%
    filter( substr( left.title,start = 1,stop = min(5,str_length(left.title),str_length(right.title)) ) != substr( right.title,start = 1,stop = min(5,str_length(left.title),str_length(right.title)) ) ) %>%
    arrange(desc(lift))

head(assoc_rules %>% select(left.title,left.year,right.title,right.year,support,confidence,lift),10)
```

Thre are many ideas that we can throw into the association rules. For example, we would like to look at modern movies that led users to view the older film.

```{r}
assoc_rules %>% 
    filter(left.year > 2000 & right.year < 1990) %>%
    arrange(desc(lift)) %>% 
    select(left.title,left.year,right.title,right.year,support,confidence,lift) %>% 
    head(20)
```

Many Ghibli's films and japanese animations appear here. It looks like modern japanese animations have enough power to draw viewers into their own world. In contrast, only few Disney animations top the chart, which can be because they are watched by nearly everyone, which resulted in lower lift scores. We are quite surprised to see Home on the Range led viewers back to The Little Mermaid. Critic reception for the film is quite low. May be that reminded viwers of Disney's renaissance era? Another notable exception is Inland Empire and Blue Velvet, which "Lynchian" structure in both films is discovered.

We can also incorporate movie's genres. We calculate the number of common genres among two films.

```{r}
assoc_rules$common_genre = apply(assoc_rules,1,function(x){
                            sum(as.numeric(x[6:23]) & as.numeric(x[26:43]))
                            })
```

Then, we mine for a movie that led viewers to a totally different kind of movie(common_genre = 0). We prefer modern films which span across different years.

```{r}
assoc_rules %>% filter(common_genre == 0) %>% 
    filter( abs(left.year - right.year) >= 5 & left.year > 2000 & right.year > 2000) %>%
    select(left.title,left.year,right.title,right.year,support,confidence,lift) %>%
    head(20)
```

The top rule consisted of both Korean movies. Thirst, which led to our favorite film: Old Boy, is a film that we have never seen before but its synopsis does sound really interesting to us! This displays the case where we may need to consider a rule on both direction as well.

Lastly, we can use assocation rules to recommend a potential movie. Let the Right One In is our favorite film and we would like to explore further movies based on it.

```{r}
assoc_rules %>% 
    filter(str_detect(left.title,"Let the Right One In") | str_detect(right.title,"Let the Right One In")) %>%
    select(left.title,left.year,right.title,right.year,support,confidence,lift) %>%
    head(20)

```

Thirst(again) appeared here. It's interesting to note that both films contain vampirism element, have the similar theme(as we guessed from the synopsis), and are not well known, which is reflected in the high lift score. The other three movies are significantly more popular. Their rules are not very interesting since many viewers also watch them anyway, regardless of the influence movie(reflected in the considerably lower lift scores). Donnie Darko, a surreal and mind-bending film, does make a bit surprise, as we didn't expect it to be heard of by so many viewers, but this perhaps reflects an enthusiasm(and bias) in the film rating communities. Also noted that the number of movies that we can recommend depended on the cutoff support value that we set. If we set this value to be too high, we will not be able to suggest anything.

# Conclusion
We apply a traditional Market Basket Analysis technique to a film recommendation setting. The technique does not provide a recommendation in a fine-grained user level, as it can be typically done by Collaborative Filtering, but it does enable us to investigate an underlying relationship within the movies. We can utilize such findings to construct a new marketing campaign, research customer's behavior, or make a product suggestion. The mining technique can also be deployed in many problem contexts, provided that they can be formulated by Basket-Item scenario.